---
title: "logtime: R package for logging and timing"
author: "Ihor Kovalyshyn, Andriy Muzychuk"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{logtime: R package for logging and timing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, message=FALSE, warning=FALSE, include=FALSE}
library(logtime)
```


In this article I would like to walk you through the main usecases of **logtime** package.


## log_time 


```log_time``` provides you with possibility of simple and efficient logging and timing of your code.


```log_time``` usage is 


```
log_time(message, level = 'DEBUG') %<% {
   expression
}
```


Let's say I would like to logtime generating of random numbers.

```{r log_time_1, echo=TRUE, message=FALSE, warning=FALSE}
log_time ('Generate random numbers') %<% {
    norm_dist_random_numbers <- rnorm(10000000)
    exp_dist_random_numbers <- rexp(10000000)    
    pois_dist_random_numbers <- rpois(10000000, lambda = 1)
}
```

```log_time``` evaluates an expression in calling environment.

```log_time``` uses pipe operator ```%<%``` what makes the code beautiful and readable.

By default it has level DEBUG but the package supports DEBUG, INFO, WARNING and  ERROR levels. 


## log_message

```log_message``` just prints a log with specified logging level. The usage is 

```
log_message(message, level = 'DEBUG')
```

So, let's create a log which tells that data frame is empty.

```{r log_message_1, echo=TRUE, message=FALSE, warning=FALSE}
log_message('The data frame is empty', level = 'WARNING')
```


## create_logger

Also the package gives you opportunity to create your own loggers with specific logging level and file for writing logs to. The logger has ```log_time``` and ```log_message``` attributes.

Let's create a logger called *clean_data* with default level INFO which will write all the logs to file *clean_data.log*.

```{r create_logger_1, echo=TRUE}
clean_data_logger <- create_logger(name = 'clean_data', 
                                   level = 'INFO', 
                                   file = 'clean_data.log'
                                   )
```

And let's logtime some process in data cleaning procedure.

```{r create_logger_2, echo=TRUE}
clean_data_logger$log_time ('Data cleaning step X') %<% {
    # some code
}
```

The code above writes logs to *clean_data.log* file with logging level INFO.
By default the logger's ```log_time``` and ```log_message``` have level INFO and can be changed. 


Now let's create simple log with changed level to ERROR.

```{r create_logger_3, echo=TRUE}
clean_data_logger$log_message('Something awful happened', level = 'ERROR')
```


## configure_logging


Last but not least there is ```configure_logging``` option in the package which sets threshold for printing logs and sets file for writing logs to.

The usage is:

```
configure_logging(threshold_level = "DEBUG", output_file = "")
```

So, to set the package for printing only logs with levels INFO and higher (namely INFO, WARNING and ERROR) we write:

```
configure_logging(threshold_level = 'INFO')
```

To set the package for printing logs to *log.log* file with levels WARNING and higher we write:

```
configure_logging(threshold_level = 'WARNING', output_file = 'log.log')
```
